Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-15 HinesHetlandGPR.tex
added more

\documentclass[10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{bm}
\title{Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method}
\author{Trever T. Hines and Eric A. Hetland}

% OUTLINE
% INTRODUCTION
% - Why we care about strain in the crust
%   - Knowing long term strain rates tells us where we can expect seismicity
%
% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processes. (include seasonal and spherical)
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
% - Numerical notes
%   - partition inverse and solve with CHOLMOD
% - outlier detection

@@

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard, since knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically.  It is then important to develop and improve upon methods for mapping strain in tectonically active regions because such maps could conceivably feed into seismic hazard models such as UCERF3 \citep{Field2014}.

Methods for estimating strain from geodetic data fall in one of two categories.  There are model-based approaches which assume that strain is the result of loading on faults which have a known geometry, and there are data-based approaches which make no assumptions about the source of deformation.  We will exclusively consider data-based approaches in this paper.  The classic and simplest method for estimating strain is to assume that the strain rate is constant in time and spatially uniform within subnetworks of the geodetic data.  Linear least squares is then used to find the components of the strain rate tensor for each subnetwork \citep[e.g][]{Frank1966,Prescott1976,Savage1986,Feigl1990,Murray2000}. Several algorithms have been developed to improve upon this procedure. \citet{Shen1996} and \citet{Shen2015} discuss an algorithm where, instead of using the immediately adjacent stations to calculate strain at a position, the strain is computed with a weighted average over the entire network where the weighting is smaller for more distant stations.  Another strategy is to fit a set of interpolating basis functions to the deformation field and then compute the strain from the analytical derivative of the interpolant \citep[e.g.][]{Beavan2001,Tape2009,Sandwell2016}.

The aforementioned studies have all been concerned with estimating long term strain rates. Time dependent strain would be useful for studying geophysical processes which occur over timescales of days to years such as slow slip events, postseismic relaxation, or volcanic deformation.  \citet{Ohtani2010} identified transient strain events by fitting a set of spatial wavelet basis functions to the deformation field at discrete time epochs, and a Kalman filtering strategy was used to ensure that the coefficients for each basis function varied smoothly in time. \citet{Holt2013} calculated time dependent strain by differentiating a bicubic interpolant that was fit to each epoch of a temporally smoothed deformation field.

Each of the methods described above are designed to overcome two complications that arise in estimating deformation gradients: (1) geodetic data are noisy and differentiation will only amplify the noise and (2) geodetic data are not observed on a regular grid, which prevents the use of standard finite difference methods for computing derivatives. In this paper we demonstrate that both of these complications can be elegantly handled with the Radial Basis Function-Finite Difference (RBF-FD) method.

The RBF-FD method was developed simultaneously and independently by \citet{Tolstykh2003}, \citet{Shu2003}, \citet{Cecil2004}, and \citet{Wright2006} as a computationally efficient way to solve large scale partial differential equations over irregular, multi-dimensional domains.  The RBF-FD method can be thought of as a generalization of the traditional finite difference method, where the node layout is no longer restricted to regular grids. Indeed, the RBF-FD method can be used to estimate derivates of discrete data located at arbitrary scattered positions in multi-dimensional space.  The RBF-FD method is particularly appealing because it is algorithmically simple, regardless of the domain shape or node layout, and also because the method has performed well in numerous benchmark tests \citep[and references therein]{Fornberg2015}.

In this paper, we do not use the RBF-FD method to solve a partial differential equation, but rather we use it to spatially smooth geodetic data and to compute deformation gradients. Our smoothing strategy can be viewed as a non-parametric, low-pass filter for scattered data where the degree of smoothness is controlled by a user specifies cutoff frequency. This can be contrasted with interpolation based methods where the resulting interpolant can be largely and perhaps unpredictably controlled by the choice of basis function. This process of smoothing and differentiating can be extended to estimate time dependent strain rates.  In that case, we first temporally smooth and differentiate GPS displacement time series to get time dependent velocities.  We then spatially smooth and differentiate the resulting velocities for each time epoch to get time dependent strain rates.

The method proposed in this paper has numerous advantages which set it apart from other methods for computing strain rates. The method is computationally efficient and stable (there is no inversion of an ill-conditioned matrix).  There are no hyper parameters or penalty parameters that need to be tuned for each application.  As opposed to interpolation strategies such as \citet{Beavan2001}, \citet{Tape2009}, or \citet{Ohtani2010}, our method assumes that velocities are locally rather than globally continuous, which allows us to easily handle discontinuities resulting from, for example, a creeping fault.

We begin this paper by summarizing the RBF-FD scheme and explaining how we construct differentiation matrices for scattered data. We then introduce the RBF-FD filter, which is used to smooth the observed geodetic data prior to differentiation.  We provide two real world demonstraitions of our method for calculating strain rates.  First we calculate the long term strain rates in Southern California from the CMM3 velocity data set \citep{Shen2011}, and we verify that our results are consistent with other studies. We then calculate time dependent strain rates in Cascadia from the GPS data provided by UNAVCO.  In Cascadia, we analyze strain resulting from slow slip events and compare it to the long term tectonic strain accumulation. Slow slip events are found to produce compression in the Olympic Peninsula, which is in addition to the compression resulting from tectonic loading.  Further south in Oregon, the slow slip events tend to release the compressional strain that is accumulated tectonically.  While similar conclusions have been drawn from fault slip inversions for slow slip events, it is important to recognize that slip inversion are the product of inverting an ill-conditioned matrix making it difficult to determine whether slip inferences are real or just an artifact of the inversion.  The strain rates presented in this paper are more direct observations and can be interpretted with a higher degree of confidence.

The Southern California Earthquake Center (SCEC) community has recently shown interest in developing methods for detecting transient signals. SCEC supported a transient detection exercise, where several research groups tested their transient detection methods with a synthetic GNSS dataset.  supported the transient Our motivation for detecting SSEs is, in part, because there have been several instances where SSEs immediately preceded megathrust earthquakes \citep{Roeloffs2006}.

GPR is a bayesian approach which does not rely on dubious parametric assumptions.

\section{Method}\label{sec:Method}
In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(z)$, where $z$ is the ordered pair $(x,t)$, $x$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(z)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\bm{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(z) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(z,z')$ is a covariance function indicating how we expect $u_i(z)$ to covary with $u_i(z')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = X(x,x')T(t,t'),
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

@@

We now consider the column vector of $n$ GNSS observations, $\bm{d}_*$. Let $\bm{y}$ be the set of $(x_i, t_j)$ pairs describing where and when each of the GNSS observations have been made. Let $\bm{a}$ be the vector of coefficients from eq. (\ref{eq:Data}) for each of the $m$ GNSS stations. We use $\bm{P}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\bm{y}$. We  also denote the vector of uncorrelated noise for each observation as $\bm{w}$, whose standard deviations are given by the formal data uncertainty $\bm{\sigma}$. The observations $\bm{d}_*$ can then be viewed as a realization of the random vector
\begin{equation}
\bm{d} = u(\bm{y}) + v(\bm{y}) + \bm{w} + \bm{P}\bm{a},
\end{equation}
which is distributed as $\mathcal{N}(\bm{0},\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T)$, where
\begin{equation}\label{eq:Cd}
\bm{\Sigma} = C_u(\bm{y},\bm{y}) + C_v(\bm{y},\bm{y}) + 
              \mathrm{diag}\left(\bm{\sigma}^2\right).  
\end{equation}
It should be understood that the notation $f(\bm{x})$ and $f(\bm{x},\bm{y})$ is shorthand for $[f(x_i)]_{x_i \in \bm{x}}$ and $[f(x_i,y_j)]_{(x_i,y_j) \in \bm{x} \times \bm{y}}$, respectively.

We now describe how to condition the prior for each component of transient displacements, $u$, with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_v$ have already been chosen. If $\kappa$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(z) = C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(z,z') = C_u(z,z') - C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}C_u(\bm{y},z').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

@@

\subsection{Outlier detection}\label{sec:Outlier}
In our formulation for estimating transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS displacement timeseries which are prone to more outliers than would be predicted for normally distributed noise. It follows that proposed methods for analyzing GNSS timeseries should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no temporally correlated noise (i.e., $v = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{y}) + \bm{P}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section X. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries.

@@

It should be noted that this algorithm does not identify jumps in GNSS time series, which are another common defect. Some, but not all, jumps can be automatically removed by parsing station logs and identifying dates of antenna changes. However, it is still necessary to manually inspect and remove jumps of unknown origin. That being said, this outlier detection algorithm significantly reduces the effort needed to manually clean GNSS data.

\section{Application to Cascadia Slow Slip Events}\label{sec:Cascadia}
In this section we estimate transient strain from GNSS data in Cascadia. In particular, we are interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. Our results illuminate where elastic strain energy, which can eventually be released seismically, is accumulating during SSEs. The estimated strain rates also reveal how strain from SSEs contributes to strain accumulated over seismic cycles. We demonstrate that transient strain rates estimated with the method described above are an effective tool for detecting SSEs and potentially other geophysical signal in GNSS data.  In Cascadia, SSEs can be detected by monitoring for associated seismic tremor \citep{Rogers2003}, which is actively being done by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. We verify that transient strain rates estimated with the method described above are indeed identifying strain from SSEs by comparing $\dot{\bm{\varepsilon}}$ to the tremor records.  We envision that GNSS derived transient strain rates could be useful for monitoring subduction zones which have SSEs but no associated tremor \citep{Schwartz2007}.

We use publicly available continuous GNSS data from the University Navstar Consortium (UNAVCO) \citep{Herring2016}, which can be found at www.unavco.org. We limit the dataset to the stations and time ranges which are pertinent to seven of the most recent SSEs in the Puget Sound region; the earliest SSE began in the August 2010, and the most recent SSE began in the February of 2017. We consider these most recent SSEs because the station coverage is sufficiently dense for us to make well constrained inferences of optimal prior models.  The positions of GNSS stations used to constrain transient strain rates are shown in Figure \ref{fig:Context}.

>>>>

\documentclass[10pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{bm}
\title{Revealing Transient Strain in Geodetic Data with Gaussian Process Regression}
\author{Trever T. Hines and Eric A. Hetland}

% OUTLINE
% INTRODUCTION

% - How people have gone about estimating strain rates
%   - Draw distinction between parametric and nonparametric methods
%   - Draw distinction between model based and data based parametric methods 
%   - Talk about Kato
%
% - Another, more challenging problem is how to calculate time-dependent strain
%   - Time dependent strain rates could illuminate transient geophysical signals such as postseismic deformation, 
%	  slow slip events, or volcanic deformation. 	  
%	- Since numerous megathrust earthquakes were preceded by slow slip, it is important to detect these features.
%	- We can also use estimates of transient strain to calibrate strain meters.	
%	- People who have tackled this problem before include Ohtani and Holt.	
%	- A major deficiency with these methods is that they are parametric giving poor uncertainties. Without good 
%	  uncertainties we cannot objectively discern signal from noise and we cannot calibrate BSM.	
% - Here we propose GPR and we tackle several two major difficulties	
%   - The method is similar to an extension of Kato to the time domain. We address two debilitating difficulties that 
%     arise. We also dont do that autocovariance bullshit.  Instead we use REML.  
%	- outliers  
%	- dense matrix
%	
% - GPR can be used for more than just estimating strain and it is a powerful geophysical tool. We can use it to quantify
%   spatial noise in GPS data. (this should be a side blurb when talking about quantifying signal.   
%
% - We discuss the method, form prior with REML, and demonstrate utility in detecting SSE by comparing it to tremor.	  
%
% METHODS
% - define model
% - discuss known deficiencies with the model
% - describe covariance as a time-space separable function 
% - describe geophysically relevant gaussian processes. (include seasonal and spherical)
% - REML for determining optimal hyperparameters
% - conditioning the prior with observations
% - differentiating the Gaussian process
% - note on addition, subtraction, and scaling
% - REML
% - Numerical notes
%   - partition inverse and solve with CHOLMOD
% - outlier detection

@@

\section{Introduction}\label{sec:Introduction}
Crustal strain rates are fundamentally important quantities for assessing seismic hazard. Knowing where and how quickly strain is accumulating gives insight into where we can expect stored elastic energy to be released seismically. Consequently, secular crustal strain rates, which are estimated from GNSS data, have been used to constrain seismic hazard models such as UCERF3 \citep{Field2014}. Dense networks of continuous GNSS stations, such as the Plate Boundary Observatory (PBO), make it feasible to estimate transient crustal strain, which we consider to be any deviation from the secular strain rate.  Transient crustal strain can potentially illuminate geophysical signal that may not be immediately apparent from inspecting the GNSS displacement data. The ability to detect transient geophysical phenomena is also relevant for seismic hazard because there are instances of major earthquakes being triggered by slow slip events \citep{Roeloffs2006} and postseismic deformation \citep{Freed2001}. For these reasons, it is important to develop and improve upon methods for estimating crustal strain rates from GNSS data.

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest and traditional method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant spatial gradients) over some subnetwork of the GNSS stations \citep[e.g][]{Frank1966,Prescott1976,Savage1986,Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined from the least squares fit to the observations. The assumption that deformation gradients are spatially uniform is obviously not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996} and \citet{Shen2015} used an inverse distance weighting scheme where the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996} and \citet{Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized velocities with bicubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The choice of parameterization can be subjective. If there are too few degrees of freedom in the parameterization then estimated strain rates will be biased and the uncertainties will be underestimated. If there are too many degrees of freedom then there will not be any coherent features in the estimated strain rates. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and plate interfaces are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

The aforementioned studies are concerned with estimating secular strain rates. In recent years the Southern California Earthquake Center (SCEC) community has shown interest in developing methods for detecting transient strain. SCEC supported a transient detection exercise \citep{Lohman2013}, where several research groups tested their methods for detecting transient geophysical signals with a synthetic GNSS dataset. Among the methods tested were the Network Strain Filter (NSF) \citep{Ohtani2010} and the Network Inversion Filter (NIF) \citep{Segall1997}. the NSF uses a wavelet parameterization to  describe the spatial component of geophysical signal and the NIF, which is intended for detecting slow fault slip, uses the elastic dislocation Green's functions from \citet{Okada1992}. For the NSF and NIF, the time dependence of the geophysical signal is modeled as integrated Brownian motion. The method described in \citet{Holt2013} was also tested in the transient detection exercise, and they calculate strain rates using a bicubic spatial parameterization of displacements between time epochs. \citet{Holt2013} defined a detection threshold based on the strain rate magnitude, and we will demonstrate that this is indeed an effective criterion for identifying geophysical signal. One challenge that is common to all these methods is that it is difficult to distinguish signal from noise, since the uncertainty on the inferred deformation can be significantly influenced by the choice of spatial parameterization.

In this paper we use Gaussian process regression (GPR) \citep{Rasmussen2006} to infer transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from discrete and potentially noisy data. GPR is closely related to Kriging \citep{Cressie1992} and Least Squares Collocation \citep{Kato1998}, and it has recently gained popularity as a tool for machine learning. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed, continuous, stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $C(t,t')=\min(t,t')$, where $t,t' \ge 0$. We use the Restricted Maximum Likelihood (REML) method to guide our choice for the prior Gaussian process.  We then incorporate the GNSS observations with the prior to form a posterior estimate of transient strain rates. The posterior transient strain rate is also a Gaussian process and, provided that our prior is sensible, has well quantified uncertainties. With well quantified uncertainties we can more confidently discern geophysical signal from noise in our estimates of transient strain rates.  We use GPR to infer transient strain rates resulting slow slip events (SSEs) in Cascadia. Our results demonstrate that GPR can be an effective tool for detecting SSEs and revealing the spatio-temporal evolution of crustal strain during SSEs.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We denote transient crustal strain rates as $\dot{\bm{\varepsilon}}(z)$, where $z$ is the ordered pair $(x,t)$, $x$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by spatially and temporally differentiating estimates of transient displacements, $\bm{u}(z)$.  As we discuss in Section \ref{sec:NoiseModel}, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$ because $\dot{\bm{\varepsilon}}$ is insensitive to common mode noise in GNSS data.  We make a prior assumption that each component of $\bm{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(z) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(z,z')$ is a covariance function indicating how we expect $u_i(z)$ to covary with $u_i(z')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = X(x,x')T(t,t'),
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

@@

We now consider the column vector of $n$ GNSS observations, $\bm{d}_*$. Let $\bm{y}$ be the set of $(x_i, t_j)$ pairs describing where and when each of the GNSS observations have been made. Let $\bm{a}$ be the vector of coefficients from eq. (\ref{eq:Data}) for each of the $m$ GNSS stations. We use $\bm{P}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\bm{y}$. We  also denote the vector of uncorrelated noise for each observation as $\bm{w}$, whose standard deviations are given by the formal data uncertainty, $\bm{\sigma}$. The observations can then be viewed as a realization of the random vector
\begin{equation}
\bm{d} = u(\bm{y}) + v(\bm{y}) + \bm{w} + \bm{P}\bm{a},
\end{equation}
which is distributed as $\mathcal{N}(\bm{0},\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T)$, where
\begin{equation}\label{eq:Cd}
\bm{\Sigma} = C_u(\bm{y},\bm{y}) + C_v(\bm{y},\bm{y}) + 
              \mathrm{diag}\left(\bm{\sigma}^2\right).  
\end{equation}
It should be understood that the notation $f(\bm{x})$ and $f(\bm{x},\bm{y})$ is shorthand for $[f(x_i)]_{x_i \in \bm{x}}$ and $[f(x_i,y_j)]_{(x_i,y_j) \in \bm{x} \times \bm{y}}$, respectively.

The prior for each component of transient displacements is then conditioned with $\bm{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \bm{d}_*$. We will assume that an appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_v$ have already been chosen. If $\kappa$ is kept finite, then we can follow \citet{Rasmussen2006} to find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(z) = C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}\bm{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(z,z') = C_u(z,z') - C_u(z,\bm{y})\left(\bm{\Sigma} + \kappa^2\bm{P}\bm{P}^T\right)^{-1}C_u(\bm{y},z').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We can use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & -\kappa^{-2} \bm{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(z) = \left[\begin{array}{cc}
                         C_u(z,\bm{y}) & \bm{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \bm{\Sigma} & \bm{P} \\
                         \bm{P}^T  & \bm{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \bm{d}_* \\
                         \bm{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(z,z') = C_u(z,z') - 
                    \left[\begin{array}{cc}
                          C_u(z,\bm{y}) & \bm{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \bm{\Sigma} & \bm{P} \\
                          \bm{P}^T  & \bm{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\bm{y},z') \\
                          \bm{0} \\
                          \end{array}\right].
\end{equation}

@@

\section{Outlier detection}\label{sec:Outlier}
In our formulation for estimating transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS displacement timeseries which are prone to more outliers than would be predicted for normally distributed noise. It follows that proposed methods for analyzing GNSS timeseries should be robust against outliers \citep[e.g.,][]{Blewitt2016}. In order to make our estimates of transient strain more robust, we automatically identify and remove outliers in the GNSS data as a pre-processing step.

Our method for detecting outliers is based on the data editing algorithm described in \citet{Gibbs2011}. We calculate the residuals between the observations and a best fitting model, and data with residuals that are anomalously large are identified as outliers. We treat $\bm{d}_*$ as a sample of $\bm{d}$ and assume that there is no temporally correlated noise (i.e., $v = 0$).  The best fitting model for $\bm{d}_*$ is considered to be the expected value of the random vector $u(\bm{y}) + \bm{P}\bm{a}$ after conditioning it with non-outlier observations.  We still consider $u$ to have a separable covariance function as in eq. (\ref{eq:TransientCovariance}), and the choice for $X$ and $T$ does not need to be the same as that used in Section \ref{sec:Method}. Since outliers are determined based on how well a spatially and temporally dependent model fits the data, we are able to identify anomalous observations which may not be immediately apparent based on inspection of individual timeseries.

@@

It should be noted that this algorithm does not identify jumps in GNSS time series, which are another common defect. Some, but not all, jumps can be automatically removed by parsing station logs and identifying dates of equipment changes. However, it is still necessary to manually find and remove jumps of unknown origin. That being said, this outlier detection algorithm significantly reduces the effort needed to manually clean GNSS data.

\section{Application to Cascadia Slow Slip Events}\label{sec:Cascadia}
We demonstrate our method for estimating transient strain rates with GNSS data from Cascadia. We are interested in identifying transient strain resulting from slow slip events (SSEs) \citep[e.g.,][]{Dragert2001}. Our results illuminate where elastic strain energy, which can eventually be released seismically, is accumulating during SSEs. We demonstrate that transient strain rates estimated with the method described above are an effective tool for detecting SSEs and potentially other geophysical signal in GNSS data.  In Cascadia, SSEs can be detected by monitoring for associated seismic tremor \citep{Rogers2003}, which is actively being done by the Pacific Northwest Seismic Network (PNSN) \citep{Wech2010}. We verify that transient strain rates estimated with the method described above are indeed identifying strain from SSEs by comparing $\dot{\bm{\varepsilon}}$ to the tremor records.  We envision that GNSS derived transient strain rates could be useful for monitoring subduction zones which have SSEs but no associated tremor \citep{Schwartz2007}.

We use publicly available continuous GNSS data from the University Navstar Consortium (UNAVCO) \citep{Herring2016}, which can be found at www.unavco.org. We limit the dataset to the stations and time ranges which are pertinent to seven of the most recent SSEs in the Puget Sound region. The earliest SSE considered began in August 2010, and the most recent SSE began in February 2017. We consider these most recent SSEs because the station coverage is sufficiently dense for us to make well constrained inferences of optimal prior models.  The positions of GNSS stations used to constrain transient strain rates are shown in Figure \ref{fig:Context}.
